Phoneme surprisal

### 2019-06-19 surprisal/phoneme_surprisal.Rmd
prelim POS tag surprisal results on Prov, progress on phoneme surprisal

---
title: "Phoneme surprisal"
author: "Josef Klafka & Dan Yurovsky"
date: "6/13/2019"
output: pdf_document
---

@@

adult_utterances <- prov_utterances %>%
  filter(speaker_role != "Target_Child") %>% 
  mutate(utterance_id = 1:n()) %>% 
  unnest_tokens(word, stem, token = stringr::str_split, pattern = "[ +]+")

adult_utterances %>%
  distinct(word) %>%
  # rowwise() %>%
  # mutate(ipa = get_ipa(word)) 
  pull(word) %>%
  write_lines("words.txt")

phones <- read_lines("phones.txt") %>%
  str_split(" ", simplify = F) %>%
  unlist() %>%
  unlist() %>%
  enframe(name = NULL) %>%
  filter(nchar(value) > 0)

adult_phonemes <- adult_utterances %>%
  left_join(ipa) %>%
  unnest_tokens(ipa, word, token = stringr::str_split, pattern = "_", drop = F) %>%
  group_by(utterance_id, word) %>%
  mutate(phon_order = 1:n()) %>%
  mutate(word_length = max(phon_order)) %>%
  ungroup()

phon_probs <- adult_utterances %>% 
  count(phon) %>%
  ungroup() %>%
  mutate(p = n/sum(n))

letter_unigrams <- adult_utterances %>% 
  left_join(letter_probs) %>%
  mutate(s = -log(p)) %>%
  group_by(word_length, letter_order) %>%
  summarise(mean_s = mean(s))

letter_unigrams %>%
  filter(word_length %in% 3:11) %>%
  ggplot(aes(x = letter_order, y = mean_s)) + 
    facet_wrap(~word_length) + 
    geom_point() +
    geom_line() + 
    xlab("Letter Position") + 
    ylab("Mean surprisal")
```

>>>>

---
title: "Phoneme surprisal"
author: Josef Klafka and Dan Yurovsky 
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: false
    number_sections: false
    theme: lumen
    toc_float: false
    code_folding: show 
editor_options: 
  chunk_output_type: console
---

@@

adult_utterances <- prov_utterances %>%
  filter(speaker_role != "Target_Child") %>% 
  mutate(utterance_id = 1:n()) %>% 
  unnest_tokens(word, stem, token = stringr::str_split, pattern = "[ +]|[_+]+")

system("rm words.txt phones.txt")
adult_utterances %>%
  distinct(word) %>%
  slice(1:10) %>%
  # filter(word != "", word != "tʃutʃutʃu", word != "ɛː") %>%
  # rowwise() %>%
  # mutate(ipa = get_ipa(word)) 
  pull(word) %>%
  write_lines("words.txt")

system('espeak -f words.txt --ipa=3 -v en-us -q --phonout="phones.txt"')
system("paste -d , words.txt phones.txt > joined.txt")

phones <- read_lines("joined.txt") %>%
  str_split(" ", simplify = F) %>%
  unlist() %>%
  unlist() %>%
  enframe(name = NULL) %>%
  filter(nchar(value) > 0)
phones

adult_phonemes <- adult_utterances %>%
  left_join(ipa) %>%
  unnest_tokens(ipa, word, token = stringr::str_split, pattern = "_", drop = F) %>%
  group_by(utterance_id, word) %>%
  mutate(ipa_order = 1:n()) %>%
  mutate(word_length = max(ipa_order)) %>%
  ungroup()

ipa_probs <- adult_phonemes %>% 
  count(ipa) %>%
  ungroup() %>%
  mutate(p = n/sum(n))

ipa_unigrams <- adult_utterances %>% 
  left_join(ipa_probs) %>%
  mutate(s = -log(p)) %>%
  group_by(word_length, ipa_order) %>%
  summarise(mean_s = mean(s))

ipa_unigrams %>%
  filter(word_length %in% 3:11) %>%
  ggplot(aes(x = ipa_order, y = mean_s)) + 
    facet_wrap(~word_length) + 
    geom_point() +
    geom_line() + 
    xlab("Letter Position") + 
    ylab("Mean surprisal")
```
